---
title: "Coursera - R Programming: Assignments"
author: "Christoph Schauer"
date: "06 May 2018"
output: html_document
---

## Preface

This notebook contains the code for assignments 1 and 3 of the course [R Programming](https://www.coursera.org/learn/r-programming) the [Data Science Specialization](https://www.coursera.org/specializations/jhu-data-science) by John Hopkins University on Coursera. All descriptions are quoted from the assignments, but lightly edited for brevity.


## Assignment 1

### Introduction

For this first programming assignment you will write three functions that are meant to interact with dataset that accompanies this assignment. The dataset is contained in a zip file specdata.zip that you can download from the Coursera web site.

### Explanation of the data

The zip file contains 332 comma-separated-value (CSV) files containing pollution monitoring data for fine particulate matter (PM) air pollution at 332 locations in the United States. Each file contains data from a single monitor and the ID number for each monitor is contained in the file name. Each file contains three variables:

* Date: the date of the observation in YYYY-MM-DD format (year-month-day)
* sulfate: the level of sulfate PM in the air on that date (measured in micrograms per cubic meter)
* nitrate: the level of nitrate PM in the air on that date (measured in micrograms per cubic meter)

For this programming assignment you will need to unzip this file and create the directory 'specdata'. Once you have unzipped the zip file, do not make any modifications to the files in the 'specdata' directory. In each file you'll notice that there are many days where either sulfate or nitrate (or both) are missing. This is common with air pollution monitoring data in the United States.  

### Part 1

Write a function named 'pollutantmean' that calculates the mean of a pollutant (sulfate or nitrate) across a specified list of monitors. The function 'pollutantmean' takes three arguments: 'directory', 'pollutant', and 'id'. Given a vector monitor ID numbers, 'pollutantmean' reads that monitors' particulate matter data from the directory specified in the 'directory' argument and returns the mean of the pollutant across all of the monitors, ignoring any missing values. [...] Please save your code to a file named pollutantmean.R.

```{r}
pollutantmean = function(directory = "specdata", pollutant = "sulfate", id = 1:332) {
     filelist = list.files(directory, full.names = TRUE)
     data = data.frame()
     for (i in id) {
          data <- rbind(data, read.csv(filelist[i]))
     }
  mean(data[, pollutant], na.rm = TRUE)
}
```

### Part 2

Write a function that reads a directory full of files and reports the number of completely observed cases in each data file. The function should return a data frame where the first column is the name of the file and the second column is the number of complete cases. [...] Please save your code to a file named complete.R. 

```{r}
complete = function(directory = "specdata", id = 1:332) {
     filelist = list.files(directory, full.names = TRUE)
     output = data.frame(id = integer(), no_obs = integer())
     for (i in id) {
          files_i <- read.csv(filelist[i])
          no_obs <- sum(complete.cases(files_i))
          storage <- data.frame(i, no_obs)
          output <- rbind(output, storage)
     }
     return(output)
}
```

### Part 3

Write a function that takes a directory of data files and a threshold for complete cases and calculates the correlation between sulfate and nitrate for monitor locations where the number of completely observed cases (on all variables) is greater than the threshold. The function should return a vector of correlations for the monitors that meet the threshold requirement. If no monitors meet the threshold requirement, then the function should return a numeric vector of length 0. [...] Please save your code to a file named corr.R.

```{r}
corr = function(directory = "specdata", threshold = 0) {
     
     files = list.files(directory, full.names = TRUE)
     output = vector()
     
     for (i in 1:332) {
          file_i = data.frame()
          file_i = read.csv(files[i])
          no_obs = length(which(complete.cases(file_i) == TRUE))
          
          if(threshold < no_obs) {
               storage_i = file_i[which(complete.cases(file_i) == TRUE), 2:3]
               corr_i = cor(storage_i)
               corr_i = corr_i[2,1]
               output = c(output, corr_i)
          }
     }
     return(output)
}
```

### Output examples

#### Part 1
```{r}
pollutantmean(directory = "specdata", pollutant = "sulfate", id = 10:20)
pollutantmean(pollutant = "nitrate")
```

#### Part 2
```{r}
complete(id = 61:65)
complete(id = c(4,5,1))
```

#### Part 3
```{r}
corr(threshold = 1000)
corr(threshold = 500)
```


## Assignment 3

### Introduction

The instructions for this assignment were pretty verbose, but the gist is the following: The data provided is a csv-file with mortality rates for 3 illnesses in US hospitals, and students have to write a couple of functions that read this file and query the data with a number of conditions.

### Assignment

### 2 Finding the best hospital in a state (version a)

```{r}
best = function(state, outcome) {
  
     data = read.csv("outcome-of-care-measures.csv", colClasses = "character")
     data[,11] = suppressWarnings(as.numeric(data[,11]))
     data[,17] = suppressWarnings(as.numeric(data[,17]))
     data[,23] = suppressWarnings(as.numeric(data[,23]))
     data = data.frame(data[,c(2, 7, 11, 17, 23)])
     colnames(data) = c("hospital", "state", "heart attack", "heart failure", "pneumonia")
  
     if(!state %in% data[,"state"]) {
          stop("invalid state")
     } 
     else if (!outcome %in% c("heart attack", "heart failure", "pneumonia")) {
          stop("invalid outcome")
     } 
     else {
          data = data[data$state == state,]
          data = data.frame(data$hospital, data[[outcome]])
          colnames(data) = c("hospital", "outcome")
          data = data[order(data$outcome, data$hospital),]
          colnames(data) = c("hospital", outcome)
     }
     return(head(data ,1))
}
```

### 2 Finding the best hospital in a state (version b)

```{r}
best = function(state, outcome) {
  
     data = read.csv("outcome-of-care-measures.csv", colClasses = "character")
     data[,11] = suppressWarnings(as.numeric(data[,11]))
     data[,17] = suppressWarnings(as.numeric(data[,17]))
     data[,23] = suppressWarnings(as.numeric(data[,23]))
     data = data.frame(data[,c(2, 7, 11, 17, 23)])
     colnames(data) = c("hospital", "state", "heart attack", "heart failure", "pneumonia")
  
     if(!state %in% data[, "state"]) {
          stop("invalid state")
     } 
     else if (!outcome %in% c("heart attack", "heart failure", "pneumonia")) {
          stop("invalid outcome")
     } 
     else {
          data = data[data$state == state, ]
          hospital = data[, "hospital"]
          outcome = data[[outcome]]
          min_value = min(outcome, na.rm = TRUE)
          output = hospital[which(outcome == min_value)]
  }
  return(output)
}
```

### 3 Ranking hospitals by outcome in a state

```{r}
rankhospital = function(state, outcome, num) {
  
     data = read.csv("outcome-of-care-measures.csv", colClasses = "character")
     data[,11] = suppressWarnings(as.numeric(data[,11]))
     data[,17] = suppressWarnings(as.numeric(data[,17]))
     data[,23] = suppressWarnings(as.numeric(data[,23]))
     data = data.frame(data[,c(2, 7, 11, 17, 23)])
     colnames(data) = c("hospital", "state", "heart attack", "heart failure", "pneumonia")
  
     if(!state %in% data[,"state"]) {
          stop("invalid state")
     } 
     else if (!outcome %in% c("heart attack", "heart failure", "pneumonia")) {
          stop("invalid outcome")
     } 
     else if (is.numeric(num)) {
          data = data[data$state == state,]
          data = data.frame(data$hospital, data[[outcome]])
          colnames(data) = c("hospital", "outcome")
          data = data[order(data$outcome, data$hospital),]
          colnames(data) = c("hospital", outcome)
          output = data[num,]
     }
     else if (!is.numeric(num)) {
          if (num == "best") {
               data = data[data$state == state,]
               data = data.frame(data$hospital, data[[outcome]])
               colnames(data) = c("hospital", "outcome")
               data = data[order(data$outcome, data$hospital),]
               colnames(data) = c("hospital", outcome)
               output = data[1,]
          }
          else if (num == "worst") {
               data = data[data$state == state,]
               data = data.frame(data$hospital, data[[outcome]])
               colnames(data) = c("hospital", "outcome")
               data = data[order(data$outcome, data$hospital, decreasing = TRUE),]
               colnames(data) = c("hospital", outcome)
               output = data[1,]
          }  
          else {
               stop("invalid num")
          }
     }
}
```

### 4 Ranking hospitals in all states

```{r}
rankall = function(outcome, num) {
     data = read.csv("outcome-of-care-measures.csv", colClasses = "character")
     data[,11] = suppressWarnings(as.numeric(data[,11]))
     data[,17] = suppressWarnings(as.numeric(data[,17]))
     data[,23] = suppressWarnings(as.numeric(data[,23]))
     data = data.frame(data[,c(2, 7, 11, 17, 23)])
     colnames(data) = c("hospital", "state", "heart attack", "heart failure", "pneumonia")
  
     if (!outcome %in% c("heart attack", "heart failure", "pneumonia")) {
          stop("invalid outcome")
     } 
     else {
          data = data.frame(data[,1:2], data[[outcome]])
          colnames(data) = c("hospital", "state", "outcome")
          states = unique(data$state)
          output = data.frame()
    
          if (is.numeric(num)) {
               for (i in 1:length(states)) {
                    data1 = data[data$state == states[i],]
                    colnames(data1) = c("hospital", "state", "outcome")
                    data1 = data1[order(data1$outcome, data1$hospital),]
                    colnames(data1) = c("hospital", "state", outcome)
                    x = data1[num,]  
                    output = rbind(output, x)
               }
          }
          else if (!is.numeric(num)) {
               if (num == "best") {
                    for (i in 1:length(states)) {
                         data1 = data[data$state == states[i],]
                         colnames(data1) = c("hospital", "state", "outcome")
                         data1 = data1[order(data1$outcome, data1$hospital),]
                         colnames(data1) = c("hospital", "state", outcome)
                         x = data1[1,]  
                         output = rbind(output, x)
                    }
               }
               else if (num == "worst") {
                    for (i in 1:length(states)) {
                         data1 = data[data$state == states[i],]
                         colnames(data1) = c("hospital", "state", "outcome")
                         data1 = data1[order(data1$outcome, data1$hospital, decreasing = TRUE),]
                         colnames(data1) = c("hospital", "state", outcome)
                         x = data1[1,]  
                         output = rbind(output, x)
                    }
               }
               else {
                    stop("invalid num")
               }
          }
     }
     return(output)
}  
```

### Exampe outputs of the functions

#### Function 2
```{r}
best("TX", "heart attack")
best("TX", "heart failure")
best("MD", "heart attack")
```

Examples for invalid inputs:
```{r, eval = FALSE}
best("XX", "heart attack")
best("AZ", "made-up illness")
```

#### Function 3
```{r}
rankhospital("TX", "heart failure", 4)
rankhospital("MD", "heart attack", "worst")
rankhospital("MN", "heart attack", 5000)
```

#### Function 4
```{r}
head(rankall("heart attack", 20), 10)
tail(rankall("pneumonia", "worst"), 3)
```
